---
title: "Bioinformatics Lab3"
author: "Milda Poceviciute, Fanny Karelius, Rab Nawaz Jan Sher and Saman Zahid"
date: "5 December 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In step one we make a vector of accession number to get data of relevant species to be get from Genbank, then we fetch data.

```{r}
x <- paste("AJ5345", 26:49, sep = "")
x <- c("Z73494", x)

#read a data from GenBank
sylvia.seq <- read.GenBank(x)
sylvia.seq
```
It can be seen that sequence length is not same for all species. The shortest sequence is of length 1041 that is some data is missing.


In next step we make a taxon of species containing name of each specie. 
We have updated the names data with accession number "Z73494" and "AJ534548" with respect to its name in genbank.

```{r}

taxa.sylvia <- attr(sylvia.seq, "species")
names(taxa.sylvia) <- names(sylvia.seq)

taxa.sylvia[1] <- "Sylvia_atricapilla"
taxa.sylvia[24] <- "Sylvia_abyssinica"

```

In this step trait data has been loaded which has mig.distance, mig.behaviour and geographicl range of sylvia.

```{r}

sylvia.eco <- read.table("sylvia_data.txt")
str(sylvia.eco)
rownames(sylvia.eco)

```

We have determined the multiple sequence alignment and converted it into DNAbin object. 
Then pairwise distance from DNA sequence has been determined by using the multiple sequence alignment created. Kimura(K80), Galtier and Gouy(GG95),Jukes and Cantor(JC69) and raw distance (that is proportion or the number of sites that differ between each pair of sequences) are calculated.

We have used pairwise.deletion = TRUE because we have observed that there is missing data in our sequence (i-e of length 1041). Pairwise.deletion is a logical indicating whether to delete the sites with missing data in a pairwise way.
```{r}

write.dna(sylvia.seq, file ="sylvia_seqs.fasta", format = "fasta")

sylvia_aligment <- msa("sylvia_seqs.fasta", type="dna")
sylvia.seq.ali <- as.DNAbin(sylvia_aligment)
rownames(sylvia.seq.ali) <- names(sylvia.seq)


syl.K80 <- dist.dna(sylvia.seq.ali, pairwise.deletion = TRUE)
syl.GG95 <- dist.dna(sylvia.seq.ali, model = "GG95", p = TRUE)
syl.JC69 <- dist.dna(sylvia.seq.ali, model = "JC69", p = TRUE)
syl.raw <- dist.dna(sylvia.seq.ali, model = "raw", p = TRUE)

```

Then we make the phylogenetic tree using neighbour joining tree and the distance matrix calculated above.

```{r}
nj.sylvia.K80 <- nj(syl.K80)
nj.sylvia.GG95 <- nj(syl.GG95)

dist.topo(nj.sylvia.K80, nj.sylvia.GG95) 

```

then we determine the topological distance (that is counts the number of bipartitions shared between two trees). The distance matrix used here are K80 and GG95.



Since nj tree returns an unrooted tree, we are converting into rooted tree by setting the outgroup(the group wrt which roots are formed) as "AJ534526" specie in function f. This function is used in boot.phylo to sample the phylogenetic tree. 


```{r}

f <- function(xx) root(nj(dist.dna(xx, p=TRUE)), "AJ534526")
tr <- f(sylvia.seq.ali)
## same than: tr <- root(nj.sylvia.K80, "AJ534526")

nj.boot.sylvia <- boot.phylo(tr, sylvia.seq.ali, f, 200,
                             rooted = TRUE)
nj.est <- tr
```

updaing tip lable of nj estimated to taxa.syliva tips labels and then plotting the estimated phylogenetic tree.

```{r}

nj.est$tip.label <- taxa.sylvia[tr$tip.label]
plot(nj.est, no.margin = TRUE)
nodelabels(round(nj.boot.sylvia / 200, 2), bg = "white")
add.scale.bar(length = 0.01)
write.tree(nj.est, "sylvia_nj_k80.tre")
```

The estimated proportion of each node has been shown in the tree.
The tree is then saved in file **sylvia_nj_k80.tre**


#Question 1.2

Creating different models by changing the model parameter.

```{r, message = FALSE, warning = FALSE}
load("sylvia.RData")
nj.est <- read.tree("sylvia_nj_k80.tre")
nj.est <- drop.tip(nj.est, "Chamaea_fasciata")
DF <- sylvia.eco[nj.est$tip.label, ]
table(DF$geo.range, DF$mig.behav)

syl.er <- ace(DF$geo.range, nj.est, type = "d") # default is "ER"
syl.sym <- ace(DF$geo.range, nj.est, type="d", model="SYM")
syl.ard <- ace(DF$geo.range, nj.est, type="d", model="ARD")

```

Model with Equal rates:
```{r}
syl.er
```

Symmetrical model:
```{r}
syl.sym
```


Model with different rates (no rate is same):
```{r}
syl.ard
```

```{r}

er_df <- data.frame(estimated_rate = syl.er$rates, standard_error = syl.er$se)

kable(er_df)

```

```{r}
sapply(list(syl.er, syl.sym, syl.ard), AIC)


```

```{r}
#install.packages("ade4")
#install.packages("mvMORPH")
#install.packages("mvSLOUCH")
#install.packages("adephylo")
library(ade4)
library(mvMORPH)
library(mvSLOUCH)
library(ape)
#devtools::install_github("kopperud/slouch")
#install.packages("rlang")
#install.packages("tibble")
library(tibble)
library(rlang)
```

# Questions 2

## Q2.1

In this question we analyse the data from _ade4_ package. It contains the phylogeny of 70 carnivora as reported by Diniz-Filho and Torres (2002), and provides the geographic range size and body size corresponding to them.

```{r}
data(carni70)
carni70_tree <- newick2phylog(carni70$tre)


size <- scalewt(log(carni70$tab))[,1] # scaled log body size in kg
names(size) <- row.names(carni70$tab)
yrange <- scalewt(carni70$tab[,2])# scaled geographic range in km
names(yrange) <- row.names(carni70$tab)
plot_data <- data.frame(spieces = row.names(carni70$tab), size = size, range =yrange )
```

Plots below show the body sizes (scaled log), the ranges (scaled), as well as size and range dependencies that correspond to the different spieces. There seem to be no obvious linear relation between these two traits.

```{r}


par(mfrow=c(1,1),las=3)

plot(plot_data$spieces, plot_data$size, xlab = "Spieces", ylab = "Size", pch = 19,
     col = rgb(0, 0, 0, 0.1))

plot(plot_data$spieces, plot_data$range, xlab = "Spieces", ylab = "Range", pch = 19,
     col = rgb(0, 0, 0, 0.1))

plot(plot_data$size, plot_data$range, xlab = "Size", ylab = "Range", pch = 19)


```

The plots below shows the phylogenetic tree. The first plot shows how the size trait evolved for different spieces. The white nodes represent the size below the mean while the black ones represent the body sizes above the mean. Similarly, the second tree plot shows the phylogenetic tree with respect to range.

```{r}
tre <- as.phylo(carni70_tree)
carni70_tree <- newick2phylog(carni70$tre)

par(mfrow = c(1, 1), oma = c(1, 1, 2, 1))
symbols.phylog(carni70_tree,size)
mtext("Size", side = 3, line = 0, outer = TRUE, font = 2, cex = 1.3)
par(fig = c(0, 1, .1, .9), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n")
legend("top", c("below mean", "above mean"), col = c("black", "black"), lty=c(NA, NA),pch = c(1, 1), lwd = c(1, 4), xpd = TRUE)

par(mfrow = c(1, 1), oma = c(1, 1, 2, 1))
symbols.phylog(carni70_tree,yrange)
mtext("Range", side = 3, line = 0, outer = TRUE, font = 2, cex = 1.3)
par(fig = c(0, 1, .1, .9), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n")
legend("top", c("below mean", "above mean"), col = c("black", "black"), lty=c(NA, NA),pch = c(1, 1), lwd = c(1, 4), xpd = TRUE)


```

## Q2.2

### Both traits evolve as independent Brownian motions

```{r}

fit1_size <- mvBM(tre, data = carni70$tab[,1], model="BM1")
fit1_range <- mvBM(tre, data = carni70$tab[,2], model="BM1")
```

The models above are independent BMs for the size and range traits. The models estimated two parameters each:
* Evolutionary rate matrix: is related to the variance. Hence, the trait size has much higher variance than the range trait:

```{r}
fit1_size$sigma
fit1_range$sigma
```

* Estimated ancestral states: this shows that the expected value of the body size is higher than the expected value of the range.

```{r}
fit1_size$theta
fit1_range$theta
```

### The traits evolve as a correlated Brownian motion.

```{r}

fit2 <- mvBM(tre, data = carni70$tab, model="BM1")
```

The same paremeters are estimated when both traits are assumed to have corellation between them:

```{r}
cat(" Evolutionary rate matrix:")
cat("\n")
fit2$sigma
cat(" Estimated ancestral states:")
cat("\n")
fit2$theta
```

We see that range and size has a positive covariance under this model. The variances and the ancestral states are very close to the onse estimated by the univariate models.

### Both traits evolve as independent Ornstein-Uhlenbeck processes

```{r,message=FALSE, warning=FALSE}

fit3_size <- mvOU(tre, data = carni70$tab[,1], model="OU1")
fit3_range <- mvOU(tre, data = carni70$tab[,2], model="OU1")

```

The two models estimated independent Ornstein-Uhlenbeck processes for each trait.
The parameters estimated are:

```{r}
cat("Alpha: strength of selection - low alpha favours a random change during the time step \n")
fit3_size$alpha
fit3_range$alpha
cat("Sigma: evolutionary rate matrix \n")
fit3_size$sigma
fit3_range$sigma
cat("Theta: estimated ancestral states \n")
fit3_size$theta
fit3_range$theta
```

It seems that the size trait is more likely to change over time than the range trait due to much lower alpha value of size. The sigma and theta values are very similar to the ones from the independent Brownian motion models.

### The traits evolve as a bivariate Ornstein-Uhlenbeck process

```{r}
phyltree = ape2ouch(tre)
fit5 <- ouchModel(phyltree = phyltree, data = carni70$tab)

```

The bivariate OU model estimated parameters:

```{r}
cat("A: strength of selection")
cat("\n")
fit5$FinalFound$ParamsInModel$A
cat("vY0: estimated ancestral state for range (OU part)")
cat("\n")
fit5$FinalFound$ParamsInModel$vY0
cat("Syy is the sigma value for range")
cat("\n")
fit5$FinalFound$ParamsInModel$Syy

```

### _size_ evolves as a Brownian motion and _range_ as an Ornstein-Uhlenbeck process adapting to it

```{r}

traits <-  data.frame(range=carni70$tab[,2],size=carni70$tab[,1])
row.names(traits) <- row.names(carni70$tab)


fit6 <-mvslouchModel(phyltree, traits,kY=1)
#
```

The bivariate OUBM model estimated these parameters:
  
```{r}
cat("A: strength of selection")
cat("\n")
fit6$FinalFound$ParamsInModel$A
cat("vY0: estimated ancestral state for range (OU part)")
cat("\n")
fit6$FinalFound$ParamsInModel$vY0
cat("vX0: estimated ancestral state for size (BM part)")
cat("\n")
fit6$FinalFound$ParamsInModel$vX0
cat("Syy is the sigma value for range")
cat("\n")
fit6$FinalFound$ParamsInModel$Syy
cat("Sxx is the sigma value for size")
cat("\n")
fit6$FinalFound$ParamsInModel$Sxx
```

We can see that sigma values changes considerbly in comparision to the first three models, but the estimated ancestral states remained almost the same.

### Comparison

All 5 models above tried to fit the parameters based on the given phylogenetic tree, and the observed trait values for the 70 spieces. We compare the models based on the AICc criterion: 

```{r}
cat("Comparison of the AICc values:\n")
cat("Both traits evolve as independent Brownian motions \n")
fit1_size$AICc + fit1_range$AICc
cat("The traits evolve as a correlated Brownian motion \n")
fit2$AICc
cat("Both traits evolve as independent Ornstein-Uhlenbeck processes \n")
fit3_size$AICc+fit3_range$AICc
cat("The traits evolve as a bivariate Ornstein-Uhlenbeck process \n")
fit5$`FinalFound`$ParamSummary$aic.c
cat("size evolves as a Brownian motion and range as an Ornstein-Uhlenbeck process \n")
fit6$`FinalFound`$ParamSummary$aic.c

```

To compare AICc criterion, we sum the values for the BM independent models. It seems the bivariate OUBM model is the best fit since it has the lowest AICc value.

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```






